---
title: "Group Project 2_Group16"
author: 'Group16 Zeyu Bai, Annika White, Hongze Shi, Wei Jin, Yijie Wu'
number-sections: true
format: pdf
include-in-header: 
  text: |
    \usepackage{fvextra}
    \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}
editor_options: 
  chunk_output_type: document
execute:
  echo: true
  eval: true
  warning: false
  message: false
---

```{r}
library(dplyr)
library(MASS)
```

# Data Processing

## Data Preprocessing

First, A new column 'season' is created based on the 'month' column, with the months categorized into seasons. For 'chip_status', we combined 'UNABLE TO SCAN' and 'SCAN NO CHIP' into a single category labelled 'No Chip'. Because the number of 'BIRD' and 'WILDLIFE' categories is much smaller compared to the other two, we filtered out 'BIRD' and 'WILDLIFE' and only studied 'CAT' and 'DOG'. The last line indicates that in the original dataset, the months are represented from 1 to 12, and the years are 2016 and 2017. Therefore, if the year is 2016, we use the month directly to represent time, but if the year is 2017, we represent time by adding 12 to the month.

```{r}
Animal <- read.csv("dataset16.csv")

Animal$year <- as.factor(Animal$year)
Animal$season <- ifelse(Animal$month %in% c(12, 1, 2), "Winter", Animal$month)
Animal$season <- ifelse(Animal$season %in% c(3, 4, 5), "Spring", Animal$season)
Animal$season <- ifelse(Animal$season %in% c(6, 7, 8), "Summer", Animal$season)
Animal$season <- ifelse(Animal$season %in% c(9, 10, 11), "Fall", Animal$season)
Animal$chip_status <- ifelse(Animal$chip_status == "SCAN CHIP", "Chip", "No Chip")
Animal <- subset(Animal, animal_type %in% c("CAT", "DOG"))
Animal$time <- Animal$month
Animal$time <- ifelse(Animal$year == 2017, Animal$month + 12, Animal$month)
```

# Poisson Model Fitting

```{r}
# Full Model
poisson_model1 <- glm(time_at_shelter ~ animal_type + year + intake_type + outcome_type + chip_status + month, family = poisson, data = Animal)
summary(poisson_model1)
```

```{r}
# Remove `animal_type`
poisson_model2 <- glm(time_at_shelter ~ year + intake_type + outcome_type + chip_status + month, family = poisson, data = Animal)
summary(poisson_model2)
```

```{r}
# Include `time`
poisson_model3 <- glm(time_at_shelter ~ animal_type + intake_type + outcome_type + chip_status + time, family = poisson, data = Animal)
summary(poisson_model3)
```

```{r}
# Remove `animal_type`
poisson_model4 <- glm(time_at_shelter ~ intake_type + outcome_type + chip_status + time, family = poisson, data = Animal)
summary(poisson_model4)
```

```{r}
# Step Selection for Final Model
null.model <- glm(time_at_shelter ~ 1, family = poisson, data = Animal)
step.model <- step(null.model,scope = list(lower = null.model, upper = poisson_model3), direction = "both")
summary(step.model)
```

The final model chosen in the step selection process includes 4 independent variables: intake type, outcome type, chip status, and time. The model equation is as follows:

$$
\begin{aligned}
\text{time at shelter}\  = \ &\beta_0 + \beta_1 \times \mathbb{I}_{\text{intake type: owner surrender}} + \beta_2 \times \mathbb{I}_{\text{intake type: stray}} + \beta_3 \times \mathbb{I}_{\text{outcome type: died}} \\
&+ \beta_4 \times \mathbb{I}_{\text{outcome type: euthanized}} + \beta_5 \times \mathbb{I}_{\text{outcome type: foster}} + \beta_6 \times \mathbb{I}_{\text{outcome type: returned to owner}} \\
&+ \beta_7 \times \mathbb{I}_{\text{chip status: no chip}} + \beta_8 \times \text{time}
\end{aligned}
$$

where,

$$
\begin{aligned}
\beta_0 &= \text{mean time at shelter when intake type is confiscated and outcome type is adopted and the animal has a chip} \\
\beta_1 \ \text{and} \ \beta_2 &= \text{effects of different intake types on time at shelter} \\
\beta_3 \ \text{to} \ \beta_6 &= \text{effects of different outcome types on time at shelter} \\
\beta_7 &= \text{effect of an animal having no chip on time at shelter} \\
\beta_8 &= \text{effect of how many months since January 2016 since the animal arrived at the shelter on time at shelter}
\end{aligned}
$$

#### Model Diagnostics for Poisson Regression

```{r}
# Check Assumptions
residuals <- residuals(poisson_model4)

plot(poisson_model4, which = 1)
plot(poisson_model4, which = 2)
```

The residuals against fitted values indicate the the residuals are scattered around the zero line indicating this assumptions appears valid.

From the QQ plot, it seems fairly reasonable that the normality assumption holds but there is some deviations in the tails of the distribution. 

```{r}
# Distribution Assumption
mean(Animal$time_at_shelter)
var(Animal$time_at_shelter)
```

Given that the mean is approximately 6.07 and the variance is substantially higher at 54.34, the data demonstrate overdispersion (where the variance is greater than the mean), which is a common characteristic in count data and can invalidate models that assume equal mean and variance such as the Poisson distribution. The model data distribution assumption here is most likely a negative binomial distribution.


# Negative-Binomial Model Fitting

```{r}
# Full Model
negbin_model1 <- glm.nb(time_at_shelter ~ animal_type + intake_type + outcome_type + chip_status + year + month, data = Animal)
summary(negbin_model1)
```

```{r}
# Use `season` to replace `month`
negbin_model2 <- glm.nb(time_at_shelter ~ animal_type + intake_type + outcome_type + chip_status + year + season, data = Animal)
summary(negbin_model2)
```

```{r}
# Use `time` to replace `year` & `season`
negbin_model3 <- glm.nb(time_at_shelter ~ animal_type + intake_type + outcome_type + chip_status + time, data = Animal)
summary(negbin_model3)
```

```{r}
# Remove `animal_type`
negbin_model4 <- glm.nb(time_at_shelter ~ intake_type + outcome_type + chip_status + time, data = Animal)
summary(negbin_model4)
```

```{r}
# Step Selection for Final Model
null.model <- glm.nb(time_at_shelter ~ 1, data = Animal)
step.model <- step(null.model, scope = list(lower = null.model, upper = negbin_model3), direction = "both")
summary(step.model)
```

The final negative-binomial model equation is represented as:

$$
\begin{aligned}
\text{time at shelter} \ = \ &\beta_0 + \beta_1 \times \mathbb{I}_{\text{intake type: owner surrender}} + \beta_2 \times \mathbb{I}_{\text{intake type: stray}} + \beta_3 \times \mathbb{I}_{\text{outcome type: died}} \\
&+ \beta_4 \times \mathbb{I}_{\text{outcome type: euthanized}} + \beta_5 \times \mathbb{I}_{\text{outcome type: foster}} + \beta_6 \times \mathbb{I}_{\text{outcome type: returned to owner}} \\
&+ \beta_7 \times \mathbb{I}_{\text{chip status: no chip}} + \beta_8 \times \text{time}
\end{aligned}
$$

where,

$$
\begin{aligned}
\beta_0 &= \text{mean time at shelter when intake type is confiscated and outcome type is adopted and the animal has a chip} \\
\beta_1 \ \text{and} \ \beta_2 &= \text{effects of different intake types on time at shelter} \\
\beta_3 \ \text{to} \ \beta_6 &= \text{effects of different outcome types on time at shelter} \\
\beta_7 &= \text{effect of an animal having no chip on time at shelter} \\
\beta_8 &= \text{effect of how many months since January 2016 since the animal arrived at the shelter on time at shelter}
\end{aligned}
$$

```{r}
# Check Assumptions
residuals <- residuals(negbin_model4)

plot(negbin_model4, which = 1)
plot(negbin_model4, which = 2)
```

The residuals appear randomly dispersed around the horizontal line without evident patterns, suggesting no major violations of model assumptions. However, slight curvilinear trends at lower fitted values may indicate potential nonlinear relationships or heteroscedasticity.

While the bulk of points follow the reference line, deviations in the tails suggest the residuals may have a non-normal distribution, which is not unexpected for count data modelled with a negative binomial distribution.

# Conclusion

In conclusion, for the distribution and model assumption, negative binomial model performs better. For the variable, outcome, intake types and general time trend without chips are all significantly negative impact with time at shelter.

In terms of further explorations, outcome type may be a confounding variable since once we know the outcome type, we also know how long the time spent at shelter is. We may also introduce new variables to replace the outcome type.
=======
# EDA

#### Numerical summary

```{r}
apply(Animal, 2, table)
```

```{r}
library(skimr)
skim(Animal)
```

#### Graphical summary

There are six boxplots for each explanatory variables, and two bar charts to show the relationship between the type of animal(cat or dog), the circumstances of their arrival at the shelter (intake_type), and their subsequent outcomes (outcome_type).

```{r}
library(ggplot2)
ggplot(data = Animal, aes(x = animal_type, y = time_at_shelter, fill = animal_type))+
  geom_boxplot()+
  labs(x = "animal_type", y = "time_at_shelter")+
  theme(legend.position = "none")+
  ggtitle("boxplot of time_at_shelter vs. animal_type")
```

Dogs tend to have a wider range and potentially a higher median time at the shelter compared to cats. There are also more outliers for dogs, indicating that some stay significantly longer than the median.

```{r}
ggplot(data = Animal, aes(x = season, y = time_at_shelter, fill = season))+
  geom_boxplot()+
  labs(x = "season", y = "time_at_shelter")+
  theme(legend.position = "none")+
  ggtitle("boxplot of time_at_shelter vs. season")
```

The boxplot shows varied medians across different outcomes. Adoption has a lower median time at the shelter compared to animals that were fostered or returned to the owner. Euthanized animals have a wide interquartile range, suggesting variable time frames before this outcome is reached.

```{r}
ggplot(data = Animal, aes(x = year, y = time_at_shelter, fill = year))+
  geom_boxplot()+
  labs(x = "year", y = "time_at_shelter")+
  theme(legend.position = "none")+
  ggtitle("boxplot of time_at_shelter vs. year")
```

The boxplot shows a lower median for the year 2016 compared to 2017, with fewer extreme outliers in 2016. This could imply a change in the shelter's operations or external factors affecting the length of stay.

```{r}
ggplot(data = Animal, aes(x = intake_type, y = time_at_shelter, fill = intake_type))+
  geom_boxplot()+
  labs(x = "intake_type", y = "time_at_shelter")+
  theme(legend.position = "none")+
  ggtitle("boxplot of time_at_shelter vs. intake_type")
```

Stray animals show a higher median and wider interquartile range of time spent at the shelter. Confiscated animals have the shortest stay, while owner surrenders are in between but closer to confiscated in terms of the median time at the shelter.

```{r}
ggplot(data = Animal, aes(x = outcome_type, y = time_at_shelter, fill = outcome_type))+
  geom_boxplot()+
  labs(x = "outcome_type", y = "time_at_shelter")+
  theme(legend.position = "none")+
  ggtitle("boxplot of time_at_shelter vs. outcome_type")
```

The boxplot shows varied medians across different outcomes. Adoption has a lower median time at the shelter compared to animals that were fostered or returned to the owner. Euthanized animals have a wide interquartile range, suggesting variable time frames before this outcome is reached.

```{r}
ggplot(data = Animal, aes(x = chip_status, y = time_at_shelter, fill = chip_status))+
  geom_boxplot()+
  labs(x = "chip_status", y = "time_at_shelter")+
  theme(legend.position = "none")+
  ggtitle("boxplot of time_at_shelter vs. chip_status")
```

Animals with a chip have a more compact interquartile range of days spent at the shelter and fewer outliers, suggesting they tend to stay for a shorter and more consistent period. In contrast, animals without a chip show a wider range and more outliers, indicating longer and more variable stays.

```{r}
ggplot(Animal, aes(x=time_at_shelter)) + 
  geom_density() +
  labs(title="Density Plot of Time at Shelter",
       x="Time at Shelter",
       y="Density")
```

This plot shows that a large proportion of animals have a very short stay at the shelter, with a sharp decrease in density as time increases. This suggests that most animals are not at the shelter for an extended period.

```{r}
ggplot(Animal, aes(x = outcome_type, y = time_at_shelter, fill = animal_type, colour = animal_type)) + 
  geom_bar(stat = "identity")
```

The bar plot indicates that for both cats and dogs, adoption and return to owner are the most common outcomes. However, dogs have a much higher rate of being returned to their owners, which might be related to longer stays in the shelter.

```{r}
ggplot(Animal, aes(x = intake_type, y = time_at_shelter, fill = animal_type, colour = animal_type)) + 
  geom_bar(stat = "identity")
```

The barchart compares the time_at_shelter by cats and dogs, broken down by intake_type. The intake types shown are "Confiscated", "Owner Surrender", and "Stray". It appears that the majority of animals in the shelter are strays, and among these, dogs tend to stay longer in the shelter than cats. The "Owner Surrender" category has a more even distribution between cats and dogs, but again, dogs show a longer shelter time overall. The "Confiscated" category has the least number of animals, but similar to the other categories, dogs have a longer shelter time than cats.

